---
title: Working with factors in R
editor: { render-on-save: true }
---

# Scope

In this article, we will cover about factors and use of forcats package.

We won’t get into the depths of where factors can be useful. However, they are mostly useful when you have a categorical variable in your data and want to use it in your visualizations or statistical model packages.

# Introduction 

Why do we want to learn how to work with factors in R?

- Mainly because you often need to work with categorical variables in R. Many functions in statistical and visualization packages in R require the input to be a factor.

# Getting started

Let's start by creating a factor vector named `foo`. 

```{r}
foo <- factor(c("green", "green", "red", "red", "yellow", "yellow", "red", "red","yellow"))
foo
```

Let's explore the factor vector that we just created. 

```{r}
table(foo)
```

The `table` function will give you the frequency of each level in the factor variable. 

Let's count number of levels in the factor variable. 

```{r}
nlevels(foo)
```

Let's see what are the levels in the factor variable.

```{r}
levels(foo)
```

If you notice from the output of the `levels` function, the levels are ordered alphabetically( this is the default behaviour). 

Let's reorder the levels in the factor variable.

```{r}
foo <- factor(foo, levels = c("yellow", "green", "red"))
levels(foo)
```

Here levels are ordered in the order that we specified. 

We can also use the `fct_relevel` function from the `forcats` package to reorder the levels based on the frequency of the levels. 

```{r}
library(forcats)
```

```{r}
foo <- fct_relevel(foo, c("yellow","red"))
levels(foo)
```

We can also change the level order based on the frequency of the levels. 

Below the levels are ordered based on the frequency of the levels, with the most frequent level first ie "red", then "yellow" and then "green".

```{r}
foo <- fct_infreq(foo)
levels(foo)
```

If we want to order it based on the frequency in the reverse order, we can do that too. 

```{r}
foo <- fct_rev(foo)
levels(foo)
```

# Word of caution

You need to be careful when working with factors in R. Even though a factor may look like a character vector, it is not one. Many functions that expect the input to be a character vector will throw an error if given a factor instead. Here is an example...

```{r}
# nchar(foo)
```

The above code will throw an error. Because the `nchar()` expects the input to be a character vector. Let's try it after coercing the factor to a character vector. 

```{r}
nchar(as.character(foo))
```

The `stringr` package functions will by default coerce the factor to a character vector and work on it. And returns the output as a character vector. But it is always better to convert the factor to a character vector before passing it to a function that expects a character vector. That way you know what you are doing, and no surprises when you are using the wrangled data in your vizualizations or statistical models. 

Here is a quick demo of how `stringr` functions work with factors.

```{r}
library(stringr)
foo_after_stringr <- str_replace(foo, "yellow", "blue")
foo_after_stringr
typeof(foo_after_stringr)
```
As you can see `foo_after_stringr` is a character vector.